---
title: "Symbolic Differentiation of R Code with Deriv Package"
subtitle: "RUG meeting in Toulouse"
author: "Serguei Sokol, IR INRA / UMR LISBP"
date: "25 septembre 2018"
output:
  beamer_presentation:
    toc: true
knit: my_render
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, error=TRUE, prompt=TRUE, size="small")
def.chunk.hook = knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x = def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)},
  document = function(x) gsub("```\n+```\n", "", x))
# in .Rprofile
# my_render <- function(input, encoding) rmarkdown::render(input, clean = FALSE, encoding = encoding)

```

# Introduction

## Existent solutions for symbolic computation

Exact and economic derivative calculation is useful e.g. for optimization codes.

Symbolic computation (including differentiation) is proposed for a long time in various maths software:

 - MAPPLE
 - Matematica
 - SAGE
 - MATLAB (dedicated toolbox)
 - Python (SymPy)
 - ...
 
In R, we have `stats::deriv()` and various packages `rSymPy`, `SymEngin.R` (not on CRAN), `Ryacas` (no more available)

So why to develop a new package `Deriv`?

## Why a new package `Deriv`?

 - Symbolic computation packages often operate on their own data type (not R's variable), e.g.
 
```{r}
library(rSymPy)
x = Var("x")
sympy("diff(sin(2*x), x, 1)")

```

## Why a new package `Deriv`?

 - `deriv()` can differentiate R expression e.g.
 
```{r}
deriv(~sin(2*x), "x")
```

but ...

## Why a new package `Deriv`?

 - ... but not a user's function body:
```{r}
y = function(x) sin(2*x)
deriv(~y(x), "x")
```
while `Deriv()` can inspect users' functions:
```{r echo=TRUE, warning=FALSE, message=FALSE, prompt=TRUE, error=TRUE}
library(Deriv)
Deriv(~y(x), "x")
```

## Why a new package `Deriv`?

 - `deriv()` return an expression that user must integrate in its code while `Deriv()` return a function that can be called
```{r}
yd = Deriv(y, "x")
class(yd)
yd(pi/2.)
```

## Why a new package `Deriv`?

 - `Deriv()` rules' table is quite complete but can be extended by user's custom derivative rules
```{r}
deriv(~besselI(x, 1), "x")
Deriv(~besselI(x, 1), "x")
```


 - `Deriv()` offers many other useful features for R programmers. We will see some of them today.

# R introspection tools

## R code as data

R code can be presented and manipulated as R data structures.
```{r size="footnotesize", tidy=TRUE}
e = quote(sin(2*x))
class(e)
(eli = as.list(e))
lapply(eli, class)
```

## R code as data

```{r}
eli[[1]] = as.symbol("cos")
(e = as.call(eli))

```
Few of functions for code manipulation:
```{r}
substitute(sin(2*x), as.environment(list(x=as.symbol("z"))))
substitute(e, as.environment(list(x=as.symbol("z"))))
do.call(substitute, list(e, as.environment(list(x=as.symbol("z")))))
```

## R code as data

 - we can explore an R function with `args()` and `body()`;
 - create functions "on the fly" with `as.function()`
 - and to know what each parameter became in a real call

```{r eval=TRUE, size="scriptsize"}
args("*")
body(y)
as.function(alist(a = , b = 2, a+b))
as.list(match.call(args("*"), quote(2*x)))
```


# `Deriv` insides

## Derivation algorithm for last leafs in AST

 Let `e` an R language element in the last leaf of Abstract Syntax Tree (AST) and `x` is symbol by which we want to differentiate.
 
`if (e == x) 1 else 0`

## Derivative rules
Next, we'll need a table of derivative rules for calls

```{r eval=FALSE, prompt=FALSE, size="footnotesize"}
# linear functions, i.e. d(f(x))/dx == f(d(arg)/dx)
dlin=c("+", "-", "c", "t", "sum", "cbind", "rbind", "list")

# rule table
# arithmetics
drule[["*"]] <- alist(e1=e2, e2=e1)
drule[["^"]] <- alist(e1=e2*e1^(e2-1), e2=e1^e2*log(e1))
drule[["/"]] <- alist(e1=1/e2, e2=-e1/e2^2)
# log, exp, sqrt
drule[["sqrt"]] <- alist(x=0.5/sqrt(x))
drule[["log"]] <- alist(x=1/(x*log(base)), base=-log(x, base)/(base*log(base)))
...
drule[["dnorm"]] <- alist(x=-(x-mean)/sd^2*if (log) 1 else dnorm(x, mean, sd),
	mean=(x-mean)/sd^2*if (log) 1 else dnorm(x, mean, sd),
	sd=(((x - mean)/sd)^2 - 1)/sd * if (log) 1 else dnorm(x, mean, sd),
	log=NULL)
...
```

## Differentiate an expression
If a call is in `drule` table, then it will be replaced by a sum of all partial derivatives read from `drule`. Each partial derivative is multiplied by a derivative of corresponding argument by `x`.

```{r}
drule[["sin"]]
drule[["*"]]
```

| Expression | Derivative |
|------------|-------------|
| `sin(2*x)` | `cos(2*x)*d(2*x)/dx` |
| `2*x`      | `x*d(2)/dx + 2*d(x)/dx` |


# Examples

## Calculate a Hessian for least squares functional

$$ L=\Sigma_i(a*x_i-b-y_i)^2 $$
   
$$   \min_{a,b} L $$
```{r}
L=function(a,b, x=NULL,y=NULL) sum((a*x+b-y)**2)
gradL=Deriv(L, c("a", "b"))
hessL=Deriv(gradL, c("a", "b"), combine="cbind")
set.seed(7); a=1; b=2; x=1:10; y=a*x+b+rnorm(x, 0, 0.1)
(sol=solve(hessL(0, 0, x, y), -gradL(0, 0, x, y)))
```

**Oh-oh!**

## What's wrong with our Hessian?

```{r size="scriptsize"}
hessL
hessL(0, 0, x, y)
```
The problem is in $\partial^2L/\partial b^2$ term which is set to 2 instead of $2n$.

Why we've got a wrong term?

Because of implicite R rule of recycling scalar arguments to fit needed vector length. We have to do it explicitly!

## Re-calculate the Hessian (the good one)
```{r size="scriptsize"}
L=function(a,b, x=NULL,y=NULL) sum((a*x+rep.int(b, length(x))-y)**2)
gradL=Deriv(L, c("a", "b"))
hessL=Deriv(gradL, c("a", "b"), combine="cbind")
set.seed(7); a=1; b=2; n=10; x=1:10; y=a*x+b+rnorm(x, 0, 0.1)
(sol=solve(hessL(0, 0, x, y), -gradL(0, 0, x, y)))
hessL
```

## Taylor series for user defined functions [^1]

```{r size="tiny", fig.height=3.75, fig.width=6}
TS=function(f, x, x0, k) {
  vd=unlist(Deriv(f, "x", nderiv=0:k)(x0))
  return(sapply(x, function(xi) sum(vd*(xi-x0)**(0:k)/factorial(0:k))))
}
f = function(x) log(x) - x^4/24 + x^3 - 6*x^2 + 3*x + 20
x=seq(0, 15, length.out = 101); x0=2
curve(f(x), xlim=c(0,10), ylim=c(-30, 30), xlab="", ylab="", lwd=2)
tmp=lapply(1:4, function(nd) curve(TS(f,x,x0,nd), xlim=c(0,10), add=T, col=nd+1, lty=5))
```

[^1]: Credits: Bertrand Koebel, Louis Pasteur University Strasbourg

## User defined derivative rules
Often classical scalar derivative applies to vector arguments, term by term. But not always:
```{r size="footnotesize"}
le=function(x) log(sum(exp(x)))
Deriv(le)
# Oops...
# Let repair it by a special rule for `le()` (thx @chuanwen)
le2=function(x) {xm=max(x); log(sum(exp(x-xm)))+xm}
dle2=function(x) {e_x=exp(x-max(x)); e_x/sum(e_x)}
drule[["le2"]]=alist(x=dle2(x))
le(710:711)
le2(710:711)
Deriv(le2)(710:711)
```

## Closer to real world example
```{r size="scriptsize"}
# Upper glycolysis model taken from "CSB Lecture flux balance analysis"
#  which cites E. Klipp, Systems Biology in Practice, 2005
upglyc=function(t, co, p) {
  # provide derivative vector for ODE solving
  # flux as function of conc and kin. parameters
  nu1=p["Vmax1"]*co["ATP"]*p["Glucose"]/(1+co["ATP"]/p["Katp1"]+p["Glucose"]/p["Kglucose1"]+co["ATP"]*p["Glucose"]/(p["Katp1"]*p["Kglucose1"]))
  nu2=p["k2"]*co["ATP"]*co["Gluc6P"]
  nu3=((p["Vfmax3"]/p["Kgluc6p3"])*co["Gluc6P"]-(p["Vrmax3"]/p["Kfruc6p3"])*co["Fruc6P"])/
    (1+co["Gluc6P"]/p["Kgluc6p3"]+co["Fruc6P"]/p["Kfruc6p3"])
  nu4=p["Vmax4"]*co["Fruc6P"]**2/(p["Kfruc6p4"]*(1+p["ka"]*(co["ATP"]/co["AMP"])**2)+co["Fruc6P"]**2)
  nu5=p["k5"]*co["Fruc16P2"]
  nu6=p["k6"]*co["ADP"]
  nu7=p["k7"]*co["ATP"]
  nu8=p["k8f"]*co["ATP"]*co["AMP"]-p["k8r"]*co["ADP"]**2
  # first derivatives in time of concentration vector co (named)
  Gluc6P=nu1-nu2-nu3
  Fruc6P=nu3-nu4
  Fruc16P2=nu4-nu5
  ATP=-nu1-nu2-nu4+nu6-nu7-nu8
  ADP=nu1+nu2+nu4-nu6+nu7+2*nu8
  AMP=-nu8
  list(c(Gluc6P, Fruc6P, Fruc16P2, ATP, ADP, AMP))
}
```

## Parameter sensitivity analysis
Function `upglyc()` is written to be usable with `DeSolve` package.
```{r}
source("upper_glyco_model.inc.R")
```

## Parameter sensitivity analysis
```{r size="footnotesize"}
Deriv(upglyc, x=c(p="Katp1"))
```



## Features remained uncovered

Some useful features remained out of scope of this presentation:

 - simplifications: they make the code to look "presentable" but do have corner cases;
 - caching : it makes the code to be efficient (no re-calculated expression), can be disabled;

# Conclusions

## Conclusions

 - `Deriv` can differentiate not only standalone R expressions or formula but a real mathematical code wrapped in a function;
 - result is computationally efficient due to symbolic simplification and caching;
 - can be helpful in *accurate* parameter sensitivity analysis
 
## Questions? Comments?

Thank you for your attention!

